---
title: "Data Tidying"
author: "Paul Schuette"
date: "May 17, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(tidyr)
```

```{r}
catch_df <- read.csv(url("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1", method = "libcurl"),
                    stringsAsFactors = FALSE)
head(catch_df)
```
```{r}
unique(catch_df$notesRegCode)
```

```{r}
catch_df <- catch_df %>% 
  select(Region, Year, Chinook, Sockeye, Coho, Chum)
head(catch_df)
```


```{r}
catch_df %>% 
  select(-All, -notesRegCode) %>% 
  head()
```


```{r}
catch_df <- catch_df %>% 
  gather(species, catch, -Region, -Year)

head(catch_df)
```

```{r}
catch_df %>% 
  spread(species, catch) %>% 
  head()
```

```{r}
catch_df <- catch_df %>% 
  rename(catch_thousands = catch)

head(catch_df)
```

```{r}
atch_df <- catch_df %>%
  mutate(catch = catch_thousands * 1000)
```

```{r}
summary(catch_df)
```

```{r}
catch_integers <- as.integer(catch_df$catch_thousands)
```

```{r}
which(is.na(catch_integers))
```

```{r}
catch_df[401,]
```

```{r}
catch_df <- catch_df %>%
  mutate(catch_thousands = ifelse(catch_thousands == "I", 1, catch_thousands),
         catch_thousands = as.integer(catch_thousands))
```


```{r}
catch_df <- catch_df %>%
  mutate(catch = catch_thousands * 1000)

summary(catch_df)
```

```{r}
catch_df <- catch_df %>% 
  select(-catch_thousands)

head(catch_df)
```

As I outlined in the Introduction, dplyr lets us employ the Split-Apply-Combine strategy and this is exemplified through the use of the group_by() and summarise() functions:

```{r}
catch_df %>% 
  group_by(Region) %>%
  summarise(mean(catch))
```

Another common use of group_by() followed by summarize() is to count the number of rows in each group. We have to use a special function from dplyr, n().

```{r}
catch_df %>% 
  group_by(Region) %>%
  summarize(n = n())
```

filter() is the verb we use to filter our data.frame to rows matching some condition. It’s similar to subset() from base R.

Let’s go back to our original data.frame and do some filter()ing:

```{r}
catch_df %>% 
  filter(Region == "SSE") %>% 
  head() # head() Show just the first n (default:  6) rows
```

```{r}
catch_df %>% 
  group_by(Region) %>% 
  summarise(mean_catch = mean(catch)) %>% 
  arrange(mean_catch)
```

```{r}
catch_df %>% 
  group_by(Region) %>% 
  summarise(mean_catch = mean(catch)) %>% 
  arrange(desc(mean_catch))
```

Joins

```{r}
region_defs <- read.csv(url("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.303.1", method = "libcurl"),
                        stringsAsFactors = FALSE)
head(region_defs, 20)
```

```{r}
region_defs <- region_defs %>% 
  select(code, mgmtArea)

head(region_defs)
```

```{r}
catch_df %>% 
  group_by(Region) %>% 
  summarise(total_catch = sum(catch)) %>% 
  left_join(region_defs, by = c("Region" = "code"))
```

